---
title: "Mendelian Randomization Exercise: LDL and Coronary Artery Disease (CAD) example toy data"
author: "Tayyaba Alvi"
output:
  html_document:
  toc: true
toc_depth: 3
theme: cosmo
---
## Mendelian Randomization Exercise: LDL and Coronary Artery Disease (CAD)
This exercise aims to guide you through a practical Mendelian Randomization (MR) analysis using a pre-harmonised dataset focusing on the causal relationship between LDL cholesterol levels and Coronary Artery Disease (CAD). You will learn to perform MR analysis, identify potential assumption violations using diagnostic plots and statistical tests, and interpret the results from different MR methods.

Objective: To determine if LDL cholesterol is a causal risk factor for CAD using Mendelian Randomization.

### Setup
First, let's load the necessary R packages and the harmonised data. We will primarily use the TwoSampleMR package.
```{r message=FALSE, warning=FALSE}
library(TwoSampleMR)
library(dplyr)
library(ggplot2)
```

```{r  message=FALSE, warning=FALSE}
#Read the harmonised and clumped data
ldl_cad_fil <- readRDS("data/LDL_CAD_excercise_data.rds")

# Display the first few rows of the harmonised data to understand its structure
head(ldl_cad_fil)
```

## Part 1: Performing the Primary MR Analysis
We will start by performing the primary MR analysis using Inverse Variance Weighted (IVW).

```{r  message=FALSE, warning=FALSE}
# Perform MR analysis using specified methods
mr_results <- mr(ldl_cad_fil, method_list = c("mr_ivw"))

# Print the MR results
print(mr_results)

```

### Questions for Part 1:
Based on the mr_results table, what are the causal effect estimate (b) and p-value (pval) for LDL on CAD for IVW?

## Part 2: Assessing Heterogeneity
Heterogeneity refers to the variation in the causal estimates across different genetic variants. Significant heterogeneity can indicate violations of the core MR assumptions.

```{r  message=FALSE, warning=FALSE}
# Perform heterogeneity test
het <- mr_heterogeneity(ldl_cad_fil)

# Print heterogeneity results
print(het)

```

## Part 2: Visualizing Individual SNP Effects (Forest Plot)
The forest plot of individual SNP effects helps visualize the consistency of the causal estimates from each genetic variant.

```{r  message=FALSE, warning=FALSE}
# Calculate single SNP effects
mr_singlesnp <- mr_singlesnp(ldl_cad_fil)

# Generate and print the forest plot for single SNP effects
p_singlesnp_forest <- mr_forest_plot(mr_singlesnp)
print(p_singlesnp_forest)

```

### Questions for Part 2:
Examine the forest plot of individual SNP effects. Do all SNPs show effects in the same direction? Do the confidence intervals largely overlap?

What does the spread or consistency of these individual SNP effects tell you about potential heterogeneity in our analysis?

How might significant heterogeneity affect the validity or interpretation of the Inverse Variance Weighted (IVW) method's results?

## Part 3: Investigating Pleiotropy with the Funnel Plot
The funnel plot helps to visually inspect for directional pleiotropy, which occurs when genetic variants affect the outcome through pathways other than the exposure.

```{r  message=FALSE, warning=FALSE}
# Generate and print the funnel plot
p_funnel <- mr_funnel_plot(mr_singlesnp)
print(p_funnel)

```

### Questions for Part 3:
Describe the shape of the funnel plot. Does it appear symmetrical? What might asymmetry indicate?

Based on the visual evidence from the funnel plot, do you suspect the presence of directional pleiotropy? Why or why not?

## Part 4: Quantifying Pleiotropy with MR Egger Intercept
The MR Egger intercept test provides a statistical test for directional pleiotropy. A significant intercept suggests pleiotropy.

```{r  message=FALSE, warning=FALSE}
# Pleiotropy test
mr_pleio <- mr_pleiotropy_test(ldl_cad_fil)
print(mr_pleio)
# Generate and print the scatter plot
mr_results <- mr(ldl_cad_fil, method_list = c("mr_ivw", "mr_egger_regression"))
p_scatter <- mr_scatter_plot(mr_results, ldl_cad_fil)
print(p_scatter)

```

### Questions for Part 4:
Interpret the egger_intercept value and its corresponding p-value. What does this statistically tell you about the presence of directional pleiotropy?

Observe the lines representing the different MR methods on the scatter plot. Which line's intercept is forced through the origin, and which is allowed to deviate? What does this illustrate about the underlying assumptions of these methods regarding pleiotropy?


Given the heterogeneity and pleiotropy test results, which MR method's estimate (IVW or MR Egger do you think is most reliable in this specific scenario, and why?

## Part 5: Leave-One-Out Analysis
The leave-one-out analysis identifies individual SNPs that might be disproportionately influencing the overall causal estimate.

```{r  message=FALSE, warning=FALSE}
# Perform leave-one-out analysis
mr_loo <- mr_leaveoneout(ldl_cad_fil)

# Generate and print the forest plot for leave-one-out analysis
# This plot shows the overall MR estimate when each SNP is left out
p_loo_forest <- mr_forest_plot(mr_loo)
print(p_loo_forest)

```

### Questions for Part 5:
Examine the leave-one-out plot. Are there any individual SNPs whose removal substantially changes the overall causal effect estimate?

What could be the implications of an influential SNP in an MR analysis, and what steps might you consider if you identify one?

### Bonus Discussion Question:
 Why IVW is not significant for this case" and "If MR Egger is underpowered, would you expect weighted median to perform better in this case?".
 
```{r  message=FALSE, warning=FALSE}

# Generate and print the scatter plot
mr_results <- mr(ldl_cad_fil, method_list = c("mr_ivw", "mr_egger_regression", "mr_weighted_median"))
print(mr_results)
p_scatter <- mr_scatter_plot(mr_results, ldl_cad_fil)
print(p_scatter)

```
